---
title: "Measuring Risk for Atrial Fibrillation: Is CHARGE-AF the Key Indicator?"
author: "Eduardo Gonzalez"
date: "2026-04-21"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(plotly)
library(dplyr)
library(tidyr)

```

# Interactive Visualizations 

Column 
-----------------------------

**How well does CHARGE-AF perform in predicting atrial fibrillation (AF) over 20 years of follow up?**

```{r Creating dataset for Figure 1}

# Create dataset with the calibration info for all models (using log-transformed CHARGE-AF as input)

calib_all <- tribble(
  ~group, ~model, ~decile, ~mphat, ~km,

  # Model 1
  "log","Model1",1,0.1257503,0.10470,
  "log","Model1",2,0.1810070,0.20919,
  "log","Model1",3,0.2280117,0.21534,
  "log","Model1",4,0.2768025,0.29905,
  "log","Model1",5,0.3305939,0.33441,
  "log","Model1",6,0.3912649,0.36313,
  "log","Model1",7,0.4605514,0.50416,
  "log","Model1",8,0.5415850,0.52166,
  "log","Model1",9,0.6464542,0.63489,
  "log","Model1",10,0.8171456,0.77738,

  # Model 2
  "log","Model2",1,0.1247103,0.10311,
  "log","Model2",2,0.1798741,0.22060,
  "log","Model2",3,0.2272660,0.21146,
  "log","Model2",4,0.2753372,0.28595,
  "log","Model2",5,0.3306681,0.30613,
  "log","Model2",6,0.3924453,0.41968,
  "log","Model2",7,0.4627647,0.46378,
  "log","Model2",8,0.5449209,0.53849,
  "log","Model2",9,0.6505642,0.66000,
  "log","Model2",10,0.8191031,0.76423,

  # Model 3
  "log","Model3",1,0.1244730,0.10854,
  "log","Model3",2,0.1798480,0.21069,
  "log","Model3",3,0.2272965,0.21318,
  "log","Model3",4,0.2753521,0.28608,
  "log","Model3",5,0.3310421,0.30348,
  "log","Model3",6,0.3926385,0.42141,
  "log","Model3",7,0.4634766,0.45667,
  "log","Model3",8,0.5453182,0.55239,
  "log","Model3",9,0.6519785,0.64993,
  "log","Model3",10,0.8196932,0.75795,

  # Model 4
  "log","Model4",1,0.1247022,0.10724,
  "log","Model4",2,0.1801624,0.21771,
  "log","Model4",3,0.2271727,0.20277,
  "log","Model4",4,0.2753760,0.29137,
  "log","Model4",5,0.3309102,0.29500,
  "log","Model4",6,0.3924362,0.42397,
  "log","Model4",7,0.4626898,0.45770,
  "log","Model4",8,0.5447232,0.55175,
  "log","Model4",9,0.6513480,0.65206,
  "log","Model4",10,0.8198827,0.75285
)


# Setting factor levels and labels 
calib_all <- calib_all %>%
  mutate(
    model = factor(model,
                   levels = c("Model1","Model2","Model3","Model4"),
                   labels = c("Model 1","Model 2","Model 3","Model 4")),
    group = recode(group,
                   orig = "Original CHARGE-AF",
                   log  = "Log-transformed CHARGE-AF",
                   af5  = "5-year AF risk")
  )


```


```{r Figure 1 }

library(plotly)
df <- calib_all %>% filter(group == "Log-transformed CHARGE-AF")
model_names <- unique(df$model)

# Start plot
p <- plot_ly()

# Add traces for each model
for(i in seq_along(model_names)) {
  m <- model_names[i]
  df_model <- df %>% filter(model == m)
  p <- add_trace(
    p,
    x = df_model$mphat,
    y = df_model$km,
    type = 'scatter',
    mode = 'lines+markers',
    name = m,
    text = paste('Decile:', df_model$decile,
                 '<br>Predicted:', round(df_model$mphat,3),
                 '<br>Observed:', round(df_model$km,3)),
    hoverinfo = 'text',
    visible = TRUE
  )
}

# Add perfect calibration line (always visible)
p <- add_trace(
  p,
  x = c(0, max(df$mphat, df$km)),
  y = c(0, max(df$mphat, df$km)),
  type = 'scatter',
  mode = 'lines',
  line = list(dash = 'dash', color = 'black'),
  name = 'Perfect calibration',
  hoverinfo = 'none',
  visible = TRUE
)

# Indices of traces for toggling (1:length(models))
trace_indices <- seq_along(model_names)

# Create buttons for dropdown (one per model)
buttons <- lapply(trace_indices, function(i) {
  vis <- rep(FALSE, length(model_names) + 1) # +1 for perfect line
  vis[i] <- TRUE  # model trace visible
  vis[length(vis)] <- TRUE  # perfect line always visible
  list(
    method = "restyle",
    args = list("visible", vis),
    label = model_names[i]
  )
})

# Layout with dropdown menu
p <- layout(
  p,
  title = "Calibration of Predicted vs. Observed AF Risk (20-Year)",
  xaxis = list(title = "Predicted atrial fibrillation risk (%)"),
  yaxis = list(title = "Observed atrial fibrillation risk (%)"),
  updatemenus = list(
    list(
      type = "dropdown",
      x = 1.1,
      y = 1,
      active = 0,
      buttons = buttons
    )
  ),
  legend = list(title = list(text = "<b>Model</b>"))
)

p



```


We can see that for all models, the predicted risk matches up well with the observed AF incidence - suggesting that **CHARGE-AF is a strong metric for predicting long term AF.**

**Real-world impact:** CHARGE-AF is an effective clinical tool for predicting long-term atrial fibrillation (AF) risk. This enables high-risk individuals to be identified early and initiate preventative treatment and lifestyle changes.

Column 
----------------------------

CHARGE-AF is a tool developed to predict atrial fibrillation from data collected at one clinic visit. But **does CHARGE-AF have a differential risk implications based on sex?**

```{r Creating dataset for Figure 2}

# Creating a dataset with all my study hazard ratios
hr_data <- tribble(
  ~Sex,   ~Predictor,         ~Model, ~HR,    ~LowerCI, ~UpperCI, ~pvalue,
  # Female
  "Female", "CHARGE-AF score at visit 4",  "Model 1", 1.205, 1.188, 1.223, 0.00001,
  "Female", "CHARGE-AF score at visit 4",  "Model 2", 1.154, 1.112, 1.198, 0.00001,
  "Female", "CHARGE-AF score at visit 3",  "Model 2", 1.073, 1.018, 1.132, 0.0088,
  "Female", "CHARGE-AF score at visit 4",  "Model 3", 1.141, 1.095, 1.189, 0.00001,
  "Female", "CHARGE-AF score at visit 3",  "Model 3", 1.039, 0.968, 1.115, 0.2897,
  "Female", "CHARGE-AF score at visit 2",  "Model 3", 1.093, 0.964, 1.240, 0.1648,
  "Female", "CHARGE-AF score at visit 4",  "Model 4", 1.142, 1.096, 1.190, 0.00001,
  "Female", "CHARGE-AF score at visit 3",  "Model 4", 1.042, 0.969, 1.120, 0.2710,
  "Female", "CHARGE-AF score at visit 2",  "Model 4", 1.082, 0.975, 1.200, 0.1368,
  "Female", "CHARGE-AF score at visit 1",  "Model 4", 0.976, 0.829, 1.150, 0.7745,
  # Male
  "Male", "CHARGE-AF score at visit 4",  "Model 1", 1.122, 1.112, 1.131, 0.00001,
  "Male", "CHARGE-AF score at visit 4",  "Model 2", 1.068, 1.044, 1.093, 0.00001,
  "Male", "CHARGE-AF score at visit 3",  "Model 2", 1.094, 1.055, 1.133, 0.00001,
  "Male", "CHARGE-AF score at visit 4",  "Model 3", 1.068, 1.044, 1.093, 0.00001,
  "Male", "CHARGE-AF score at visit 3",  "Model 3", 1.052, 1.004, 1.103, 0.0351,
  "Male", "CHARGE-AF score at visit 2",  "Model 3", 1.084, 1.020, 1.152, 0.0091,
  "Male", "CHARGE-AF score at visit 4",  "Model 4", 1.059, 1.034, 1.085, 0.00001,
  "Male", "CHARGE-AF score at visit 3",  "Model 4", 1.050, 1.001, 1.102, 0.0460,
  "Male", "CHARGE-AF score at visit 2",  "Model 4", 1.073, 0.989, 1.163, 0.0913,
  "Male", "CHARGE-AF score at visit 1",  "Model 4", 1.021, 0.923, 1.130, 0.6889
)

# Let's order the factors for plotting 
hr_data <- hr_data %>%
  mutate(
    Sex = factor(Sex, levels = c("Female", "Male")),
    Model = factor(Model, levels = c("Model 1","Model 2","Model 3","Model 4"))
  )
    

```

```{r Figure 2}

# Filtering for Visit 4 only -- this is the most important risk score
visit4_data <- hr_data %>%
  filter(Predictor == "CHARGE-AF score at visit 4")

# Ensuring the my models are ordered correctly
visit4_data <- visit4_data %>%
  mutate(Model = factor(Model, levels = c("Model 1","Model 2","Model 3","Model 4")))

# Creating the plot
p <- plot_ly()

# Adding a trace for each sex to enable comparisons
for(s in levels(visit4_data$Sex)) {
  df_sex <- visit4_data %>% filter(Sex == s)
  p <- p %>%
    add_trace(
      data = df_sex,
      x = ~Model,
      y = ~HR,
      type = 'scatter',
      mode = 'markers',
      marker = list(size = 10),
      error_y = list(
        type = "data",
        array = ~UpperCI - HR,
        arrayminus = ~HR - LowerCI,
        visible = TRUE
      ),
      name = s,
      text = ~paste("Sex:", Sex,
                    "<br>Model:", Model,
                    "<br>HR:", sprintf("%.3f", HR),
                    "<br>95% CI:", sprintf("%.3f", LowerCI), "-", sprintf("%.3f", UpperCI)),
      hoverinfo = "text"
    )
}

# Adding HR = 1 (i.e., no effect) as the reference line 
p <- p %>%
  add_lines(
    x = levels(visit4_data$Model),
    y = rep(1, length(levels(visit4_data$Model))),
    line = list(dash = 'dash', color = 'gray50'),
    name = 'HR = 1',
    hoverinfo = 'none'
  )

# Adding a dropdown menu to filter by sex 
p <- p %>%
  layout(
    title = "Hazard Ratios for Atrial Fibrillation by Model",
    yaxis = list(title = "Hazard Ratio (95% CI)"),
    xaxis = list(title = "Model"),
    updatemenus = list(
      list(
        type = "dropdown",
        x = 0, y = 1.2,
        buttons = list(
          list(method = "restyle",
               args = list("visible", c(TRUE, FALSE, TRUE)), # Female only + HR=1
               label = "Females"),
          list(method = "restyle",
               args = list("visible", c(FALSE, TRUE, TRUE)), # Male only + HR=1
               label = "Males"),
          list(method = "restyle",
               args = list("visible", c(TRUE, TRUE, TRUE)),  # Both + HR=1
               label = "Both")
        )
      )
    )
  )

p


```

An increase in CHARGE-AF risk score results in a **larger increase in AF risk among females than among males.**

 
Please visit following link to access the GitHub repository for this project: https://github.com/Eduardo-GV/Thesis-Dashboard/tree/main


# Model Interpretation

Models 1 through 4 are Cox regression models that use CHARGE-AF scores to predict incident AF. Model 4 includes only one (baseline) CHARGE-AF risk score, while each subsequent model incorporates an additional (prior) risk score. The calibration and hazard ratios produced by all models are very similar – suggesting that **one risk score alone is sufficient to ascertain AF risk.** Incorporating prior risk scores does not contribute additional meaningful information. 


# Data Description 

Data (not publicly-available) is taken from the ARIC study, which followed a cohort of 15,792 adults beginning in 1987. Participants of ages 45 to 64 were recruited from four U.S communities: Forsyth County, North Carolina; Jackson, Mississippi; suburban Minneapolis, Minnesota; and Washington County, Maryland. My final analytic sample (N=10,188) includes participants with **complete measurements from visits 1 through 4** of the ARIC study (Visit 1: 1987–89; Visit 2: 1990–92; Visit 3: 1993–95; Visit 4: 1996–98). Data collection includes measurements obtained during clinic visits and through telephone interviews. Participants were **excluded if they had prevalent AF detected at or before visit 4.** 


